A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-11-05 latex/main.tex
Updates from Overleaf

\subsection*{Comparison with UShER+RIPPLES}

Things to say.
\begin{itemize}
\item Explain how it works. Post-hoc. Note methodological distinction 
with sc2ts.
\item Comparison at k/p=4 overall.
\item Discussion of X lineages.
\item Discussion of parameter choice and our lower bound on recomb.
\end{itemize}

The most comprehensive analysis on recombination in SARS-CoV-2 to 
date has been performed using the combination of UShER and 
RIPPLES~\cite{Turakhia2022}. RIPPLES works by taking a
pre-existing UShER phylogeny and systematically breaking up
sequences subtended by long branches into multiple segments 
that are placed independently.
A recombination event at a node is flagged if
the overall parsimony of these multiple placements
sufficiently improves over
the original single-parent origin in the phylogeny.
This approach has much in common with the LS HMM
used in sc2ts, but with one important distinction.
While sc2ts detects and incorporates recombination
in ``real-time'' as sequences are added to the ARG in daily
batches, RIPPLES operates on a complete phylogeny,
attempting to detect recombination post-hoc among millions of nodes. 
Like any phylogenetic reconstruction method
that ignores recombination, UShER must force placement of recombinant
sequences in places that minimize the discrepancy across the entire sequence,
which may not be near any of the parent lineages. 
These permanent misplacements distort the phylogeny~\cite{Schierup2000consequences}
and subsequent tree reorganisations to maximise overall parsimony~\cite{DeMaio2023} 
may break up the erroneous long branches used to detect 
recombination.

[Text from Sally]
The main advantage of sc2ts over previous methods to infer recombination
phylogenetically is that recombination events are explicitly handled as the ARG
is built. Other methods (e.g., RIPPLES) first build a phylogeny, assuming no
recombination, and then look for unusual features (e.g., unusually long
branches). Phylogenetic methods that ignore recombination, however, distort the
tree shape (REFS - Schierup and Hein 2000) and force placement of recombinant
lineages in places that minimize the discrepancy across the entire sequence,
rather than near either of the parent lineages.  As a consequence, many true
recombination events can be missed because their branches are artificially
inferred to be short on reconstructed trees that do not incorporate
recombination. In addition, many of the recombinant sequences can be placed
before their parents in time, because recombinants can appear to be more
similar to ancestral sequences that are known with more uncertainty [OR because
recombinants are placed anywhere on a tree that best fits].

This ``parsimony improvement'' parameter ($p$)
is equivalent to the sc2ts recombination penalty ($k$).

RIPPLES~\cite{Turakhia2022} is a method for identifying
recombination events using a phylogeny inferred by UShER.
This distinguishes RIPPLES from other recombination detection
methods such as 3SEQ, which classify particular \emph{sequences}
as recombinant or not and can only identify 
potential onward transmission post-hoc. 
The combination of UShER and RIPPLES
is therefore the only method that is directly comparable to sc2ts. 
RIPPLES works by first identifying long 
branches in the input tree, by default those carrying 3 or more 
mutations. For each candidate branch,
RIPPLES systematically splits the sequence into 2 or 3 segments, and
finds parsimonious placements for each segment independently.

@@

that from UShER+RIPPLES,
largely reflecting differences in model formulation and parameter choice.
When applied \textit{post hoc} to an UShER phylogeny of
$\sim$1.6 million genomes sampled up to May, 2021,
RIPPLES reported 589 unique recombination events with 43,104 descendant
samples--corresponding to $\sim$2.7\% of all samples of recombinant origin.
Using the same time interval, sc2ts \(k=4\) identifies
14 high-confidence recombination events with
172 descendant samples among 501,235 samples \(0.034\%\),
a 40- to 80-fold reduction.
Much of this difference is explained by
the stricter parsimony penalty used here:
increasing the penalty from p=3 (as used in the RIPPLES analysis) to
k=4 greatly reduces false positives
while retaining nearly all verified recombinant lineages,
including those missed by RIPPLES at equivalent thresholds.
These indirect comparisons suggest that
our estimate represents a conservative lower bound.
In future studies, dissecting the methodological differences
between sc2ts and UShER+RIPPLES may
inspire new ideas to develop or improve ARG building methods
for SARS-CoV-2 (and potentially other pathogens).

>>>>

\subsection*{Comparison with UShER+RIPPLES}
RIPPLES~\cite{Turakhia2022} is a method for identifying
recombination events using a phylogeny inferred by UShER.
This distinguishes RIPPLES from other recombination detection
methods such as 3SEQ, which classify particular \emph{sequences}
as recombinant or not and can only identify 
potential onward transmission post-hoc. 
The combination of UShER and RIPPLES
is therefore the only method that is directly comparable to sc2ts. 
RIPPLES works by first identifying long 
branches in the input tree, by default those carrying 3 or more 
mutations. For each candidate branch,
RIPPLES systematically splits the sequence into 2 or 3 segments, and
finds parsimonious placements for each segment independently.
A recombination event at that node is then flagged if
the overall parsimony of these multiple placements
sufficiently improves over
the original single-parent origin in the phylogeny.
This ``parsimony improvement'' parameter ($p$)
is equivalent to the sc2ts recombination penalty ($k$).

@@
